Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2025-01-09 paper/main.tex
cleaned up fig 6 results section

We carried out three different versions of the analyses described above,
wherein we considered different sources of information in our estimates of
participants' knowledge for each quiz question. First, we estimated knowledge
at each question's embedding coordinate using \textit{all} other questions
answered by the same participant on the same quiz (``All questions'';
Fig.~\ref{fig:predictions}, top row). This test was intended to assess the
overall predictive power of our approach. Second, we estimated knowledge for
each question about a given lecture using only the other questions (from the
same participant and quiz) about that \textit{same} lecture
(``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows). This test was
intended to assess the \textit{specificity} of our approach by asking whether
our predictions could distinguish between questions about different content
covered by the same lecture. Third, we estimated knowledge for each question
about one lecture using only the questions (from the same participant and quiz)
about the \textit{other} lecture (``Across-lecture'';
Fig.~\ref{fig:predictions}, bottom rows). This test was intended to assess the
\textit{generalizability} of our approach by asking whether our predictions
could extend across the content areas of the two lectures. When computing these
knowledge estimates, we used a rebalancing procedure to ensure that (for a given
participant and quiz) the knowledge estimates for correctly and incorrectly
answered questions were based on the same proportion of correctly answered
questions (see~\nameref{subsec:glmm}).

%In performing these analyses, our null hypothesis is that the knowledge
%estimates we compute based on the quiz questions' embedding coordinates do
%\textit{not} provide useful information about participants' abilities to answer
%those questions. In other words, our estimates of participants' knowledge
%for questions they answered correctly should (on average) be
%
%In other words, since our knowledge estimates are computed
%as a weighted proportion-correct score (where each held-in question's weight reflects
%its embedding-space distance from the held-out question; see
%Eqn.~\ref{eqn:prop}), if these weights are uninformative (e.g., randomly
%distributed), then we should expect that, on average,
%
%% note that with correction, pcorrect of held in questions is now $\frac{n - 1}{q - 1}$ regardless of held-out question's label
%
%==================================
%
%In performing these analyses, our null hypothesis is that the knowledge
%estimates we compute based on the quiz questions' embedding coordinates do
%\textit{not} provide useful information about participants' abilities to answer
%those questions. What result might we expect to see if this is the case? To
%gain an intuition for this scenario, consider the expected outcome if we
%carried out these same analyses using a simple proportion-correct measure in
%lieu of our knowledge estimates. Suppose a participant correctly answered $n$
%out of $q$ questions on a given quiz. If we hold out a single
%\textit{correctly} answered question, the proportion of remaining questions
%answered correctly would be $\frac{n - 1}{q - 1}$. Whereas if we hold out a
%single \textit{incorrectly} answered question, the proportion of remaining
%questions answered correctly would be $\frac{n}{q - 1}$. In this way, the
%proportion of correctly answered remaining questions is always \textit{lower}
%when the held-out question was answered correctly than when it was answered
%incorrectly. Because our knowledge estimates are computed as a weighted version
%of this same proportion-correct score (where each held-in question's weight
%reflects its embedding-space distance from the held-out question; see
%Eqn.~\ref{eqn:prop}), if these weights are uninformative (e.g., randomly
%distributed), then we should expect to see this same inverse relationship
%between estimated knowledge and performance, on average. On the other hand, if
%the spatial relationships among the quiz questions' embeddings \textit{are}
%predictive of participants' knowledge about the questions' content, then we
%would expect \textit{higher} estimated knowledge for held-out correctly (versus
%incorrectly) answered questions.
%
%Before presenting our results, it is worth considering three possible
%explanations of why a participant might answer a given question correctly or
%incorrectly. One possibility is that the participant simply \textit{guessed}
%the answer. A second is that they selected the incorrect answer by mistake,
%despite ``knowing'' the correct answer (or vice versa). In both of these
%scenarios, the participant's knowledge about the question's content should be
%uninformative about their observed response. A third possibility is that the
%participant's response reflects their \textit{actual} knowledge about the
%question's content. In this case, we \textit{might} expect to see a positive
%relationship between the participant's knowledge and their likelihood of
%answering the question correctly. However, in order to see this positive
%relationship, the participant's knowledge must be structured in a way that is
%reflected (at least partially) by the embedding space. In other words, if the
%participant's performance reflects their true knowledge, but our text embedding
%space does not sufficiently capture the structure of that knowledge, then the
%knowledge estimates we generate will not be predictive of the participant's
%performance. In the extreme, if the embedding space is completely unstructured
%with respect to the content of the quiz questions, then we would expect to see
%the negative relationship between estimated knowledge and performance that we
%described above.

%%\textbf{NOTES FOR REWRITING RESULTS}
%%\begin{itemize}
%%    \item all questions
%%    \begin{itemize}
%%        \item quiz 1: odds ratio $(OR) = 8.126$, 95\%\ $\textnormal{CI} = [3.116,\ 20.123]$, likelihood-ratio test statistic $(\lambda_{LR}) = 17.002,\ p < 0.001$
%%        \item quiz 2: $OR = 14.902,\ 95\%\ \textnormal{CI} = [4.976,\ 39.807],\ \lambda_{LR} = 25.408,\ p < 0.001$
%%        \item quiz 3: $OR = 37.409,\ 95\%\ \textnormal{CI} = [10.425,\ 107.145],\ \lambda_{LR} = 40.948,\ p < 0.001$
%%        \item we can reliably predict success on individual held-out questions regardless of level of knowledge; predictive power grows as participants' knowledge space becomes better defined/structured
%%    \end{itemize}
%%    \item within-lecture
%%    \begin{itemize}
%%        \item quiz 1, lecture 1: $OR = 15.934,\ 95\%\ \textnormal{CI} = [5.173,\ 38.005],\ \lambda_{LR} = 40.971,\ p < 0.001$
%%        \item quiz 1, lecture 2: $OR = 9.775,\ 95\%\ \textnormal{CI} = [2.93,\ 25.08],\ \lambda_{LR} = 13.924,\ p < 0.001$
%%        \item quiz 2, lecture 1: $OR = 35.126,\ 95\%\ \textnormal{CI} = [5.113,\ 123.868],\ \lambda_{LR} = 32.251,\ p < 0.001$
%%        \item quiz 2, lecture 2: $OR = 4.717,\ 95\%\ \textnormal{CI} = [2.021,\ 9.844],\ \lambda_{LR} = 16.788,\ p < 0.001$
%%        \item quiz 3, lecture 1: $OR = 2.485,\ 95\%\ \textnormal{CI} = [0.724,\ 8.366],\ \lambda_{LR} = 1.984,\ p = 0.170$
%%        \item quiz 3, lecture 2: $OR = 16.902,\ 95\%\ \textnormal{CI} = [3.353,\ 53.265],\ \lambda_{LR} = 23.233,\ p < 0.001$
%%        \item can successfully distinguish between knowledge for questions about different content within a lecture, but may be susceptible to ceiling effects (considering questions included in model, across-participants avg prop correct on quiz3, lecture-1 questions was highest of any set of questions we analyzed (mean: 0.795, std: 0.236)
%%    \end{itemize}
%%    \item across-lecture
%%    \begin{itemize}
%%        \item quiz 1, lecture 1: $OR = 1.896,\ 95\%\ \textnormal{CI} = [0.419,\ 9.088],\ \lambda_{LR} = 0.712,\ p = 0.404$
%%        \item quiz 1, lecture 2: $OR = 1.522,\ 95\%\ \textnormal{CI} = [0.332,\ 6.835],\ \lambda_{LR} = 0.286,\ p = 0.611$
%%        \item quiz 2, lecture 1: $OR = 3.49,\ 95\%\ \textnormal{CI} = [0.739,\ 12.849],\ \lambda_{LR} = 3.266,\ p = 0.083$
%%        \item quiz 2, lecture 2: $OR = 2.199,\ 95\%\ \textnormal{CI} = [0.711,\ 5.623],\ \lambda_{LR} = 2.304,\ p = 0.141$
%%        \item quiz 3, lecture 1: $OR = 11.294,\ 95\%\ \textnormal{CI} = [1.375,\ 47.744],\ \lambda_{LR} = 10.396,\ p < 0.001$
%%        \item quiz 3, lecture 2: $OR = 7.302,\ 95\%\ \textnormal{CI} = [1.077,\ 44.879],\ \lambda_{LR} = 4.708,\ p = 0.038$
%%        \item ability to generalize across content areas depends on participants having structured knowledge about both (A) the content used to estimate knowledge, and (B) the content for which knowledge is estimated (quiz 1: have neither; quiz 2: have one but not the other; quiz 3: have both)
%%    \end{itemize}
%%\end{itemize}

%When we fit a GLMM to estimates of participants' knowledge for each Quiz~1
%question based on all other Quiz~1 questions, we observed an outcome consistent
%with our null hypothesis: higher estimated knowledge at the embedding
%coordinate of a held-out question was associated with a lower likelihood of
%answering the question correctly (odds ratio $(OR) = 0.136$, likelihood-ratio
%test statistic $(\lambda_{LR}) = 19.749$, 95\%\ $\textnormal{CI} = [14.352,\
%26.545],\ p < 0.001$). This outcome suggests that our knowledge estimates do
%\textit{not} provide useful information about participants' Quiz~1 performance
%when we aggregated across all question content areas. We speculated that this
%might either indicate that the knowledge estimates are uninformative in
%general, or about Quiz~1 performance in particular. This would be expected, for
%example, if participants were guessing about the answers to the Quiz~1
%questions (prior to having watched either lecture). When we repeated this
%analysis for Quizzes~2 and~3, we found that higher estimated knowledge
%for a given question predicted a \textit{greater} likelihood of answering it correctly
%(Quiz~2: $OR = 2.905,\ \lambda_{LR} = 17.333,\ 95\%\ \textnormal{CI} =
%[14.966,\ 29.309],\ p = 0.001$; Quiz~3: $OR = 3.238,\ \lambda_{LR} = 6.882,\
%95\%\ \textnormal{CI} = [6.228,\ 8.184],\ p = 0.016$). Taken together, these
%results suggest that our knowledge estimates reliably predict participants'
%performance on individual held-out quiz questions, but only after participants
%have received at least some training.

%When we fit a GLMM to estimates of participants' knowledge for each Quiz~1
%question based on all other Quiz~1 questions, we found that higher estimated
%knowledge for a given question predicted a greater likelihood of answering it
%correctly (odds ratio $(OR) = 8.126$, 95\%\ $\textnormal{CI} = [3.116,\
%20.123]$, likelihood-ratio test statistic $(\lambda_{LR}) = 17.002,\ p <
%0.001$). This relationship held and appeared to strengthen when we repeated this
%analysis for Quiz~2 ($OR = 14.902,\ 95\%\ \textnormal{CI} = [4.976,\ 39.807],\
%\lambda_{LR} = 25.408,\ p < 0.001$), and again for Quiz~3 ($OR = 37.409,\ 95\%\
%\textnormal{CI} = [10.425,\ 107.145],\ \lambda_{LR} = 40.948,\ p < 0.001$).
%Taken together, these results suggest that our knowledge estimates reliably
%predict participants' performance on individual questions when aggregated across
%all quiz content, and that their predictive power increases as participants gain
%additional knowledge about the content captured by the text embedding space.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Finally, when we estimated participants' knowledge for each question about one
lecture using their performance on questions (from the same quiz) about the
\textit{other} lecture, we observed a somewhat different pattern of results.
Here we found that before viewing either lecture (i.e., on Quiz~1),
participants' abilities to answer \textit{Four Fundamental Forces}-related
questions could not be predicted from their responses to \textit{Birth of
Stars}-related questions ($OR = 1.896,\ 95\%\ \textnormal{CI} = [0.419,\
9.088],\ \lambda_{LR} = 0.712,\ p = 0.404$), nor could their abilities to
answer \textit{Birth of Stars}-related questions be predicted from their
responses to \textit{Four Fundamental Forces}-related questions ($OR = 1.522,\
95\%\ \textnormal{CI} = [0.332,\ 6.835],\ \lambda_{LR} = 0.286,\ p = 0.611$).
We similarly found that participants' success on questions about either lecture
could not be predicted given their responses to questions about the other
lecture after viewing \textit{Four Fundamental Forces} but before viewing \textit{Birth of Stars} (i.e., on Quiz~2; \textit{Four Fundamental Forces}
questions given \textit{Birth of Stars} questions: $OR = 3.49,\ 95\%\
\textnormal{CI} = [0.739,\ 12.849],\ \lambda_{LR} = 3.266,\ p = 0.083$;
\textit{Birth of Stars} questions given \textit{Four Fundamental Forces}
questions: $OR = 2.199,\ 95\%\ \textnormal{CI} = [0.711,\ 5.623],\
\lambda_{LR} = 2.304,\ p = 0.141$). Only after viewing \textit{both} lectures
(i.e., on Quiz~3) did these across-lecture knowledge estimates reliably predict
participants' success on individual quiz questions (\textit{Four Fundamental
Forces} questions given \textit{Birth of Stars} questions: $OR = 11.294,\
95\%\ \textnormal{CI} = [1.375,\ 47.744],\ \lambda_{LR} = 10.396,\ p < 0.001$;
\textit{Birth of Stars} questions given \textit{Four Fundamental Forces}
questions: $OR = 7.302,\ 95\%\ \textnormal{CI} = [1.077,\ 44.879],\
\lambda_{LR} = 4.708,\ p = 0.038$). Taken together, these results suggest that
our knowledge estimates can be used to predict participants' success across
content areas once they have received some training on both the content about
which their knowledge is estimated, and the content used to construct these
estimates.

>>>>

We carried out three different versions of the analyses described above,
wherein we considered different sources of information in our estimates of
participants' knowledge for each quiz question. First, we estimated knowledge
at each question's embedding coordinate using \textit{all} other questions
answered by the same participant on the same quiz (``All questions'';
Fig.~\ref{fig:predictions}, top row). This test was intended to assess the
overall predictive power of our approach. Second, we estimated knowledge for
each question about a given lecture using only the other questions (from the
same participant and quiz) about that \textit{same} lecture
(``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows). This test was
intended to assess the \textit{specificity} of our approach by asking whether
our predictions could distinguish between questions about different content
covered by the same lecture. Third, we estimated knowledge for each question
about one lecture using only the questions (from the same participant and quiz)
about the \textit{other} lecture (``Across-lecture'';
Fig.~\ref{fig:predictions}, bottom rows). This test was intended to assess the
\textit{generalizability} of our approach by asking whether our predictions
could extend across the content areas of the two lectures. When computing these
knowledge estimates, we used a rebalancing procedure to ensure that (for a given
participant and quiz) the knowledge estimates for correctly and incorrectly
answered questions were computed from the same proportion of correctly answered
questions (see~\nameref{subsec:glmm}).

@@

Finally, when we estimated participants' knowledge for each question about one
lecture using their performance on questions (from the same quiz) about the
\textit{other} lecture, we observed a somewhat different pattern of results.
Here we found that before viewing either lecture (i.e., on Quiz~1),
participants' abilities to answer \textit{Four Fundamental Forces}-related
questions could not be predicted from their responses to \textit{Birth of
Stars}-related questions ($OR = 1.896,\ 95\%\ \textnormal{CI} = [0.419,\
9.088],\ \lambda_{LR} = 0.712,\ p = 0.404$), nor could their abilities to
answer \textit{Birth of Stars}-related questions be predicted from their
responses to \textit{Four Fundamental Forces}-related questions ($OR = 1.522,\
95\%\ \textnormal{CI} = [0.332,\ 6.835],\ \lambda_{LR} = 0.286,\ p = 0.611$).
We similarly found that participants' success on questions about either lecture
could not be predicted given their responses to questions about the other
lecture after viewing \textit{Four Fundamental Forces} but before viewing \textit{Birth of Stars} (i.e., on Quiz~2; \textit{Four Fundamental Forces}
questions given \textit{Birth of Stars} questions: $OR = 3.49,\ 95\%\
\textnormal{CI} = [0.739,\ 12.849],\ \lambda_{LR} = 3.266,\ p = 0.083$;
\textit{Birth of Stars} questions given \textit{Four Fundamental Forces}
questions: $OR = 2.199,\ 95\%\ \textnormal{CI} = [0.711,\ 5.623],\
\lambda_{LR} = 2.304,\ p = 0.141$). Only after viewing \textit{both} lectures
(i.e., on Quiz~3) did these across-lecture knowledge estimates reliably predict
participants' success on individual quiz questions (\textit{Four Fundamental
Forces} questions given \textit{Birth of Stars} questions: $OR = 11.294,\
95\%\ \textnormal{CI} = [1.375,\ 47.744],\ \lambda_{LR} = 10.396,\ p < 0.001$;
\textit{Birth of Stars} questions given \textit{Four Fundamental Forces}
questions: $OR = 7.302,\ 95\%\ \textnormal{CI} = [1.077,\ 44.879],\
\lambda_{LR} = 4.708,\ p = 0.038$). Taken together, these results suggest that
our knowledge estimates can be used to predict participants' success across
content areas once they have received some training on both the content about
which their knowledge is estimated and the content used to construct these
estimates.
